---
title: Algebraic perfect hash functions
author: "Alex Towell"
output:
  bookdown::html_book:
    toc: yes
  bookdown::pdf_document2:
    keep_tex: true
    extra_dependencies: ["amsmath","mathtools"]
  bookdown::pdf_book:
    toc: yes
    extra_dependencies: ["amsmath","mathtools"]
  bookdown::gitbook:
    config:
      toc:
        collapse: subsection
        scroll_highlight: true
        before: null
        after: null
      toolbar:
        position: fixed
      edit : null
      download: null
      search:
        # options to control/tune search engine behavior (for
        engine: lunr # or fuse
        # fuse.js, refer to https://fusejs.io/api/options.html)
        options: null
      fontsettings:
        theme: white
        family: sans
        size: 2
      sharing:
        facebook: true
        github: false
        twitter: true
        linkedin: false
        weibo: false
        instapaper: false
        vk: false
        whatsapp: false
        all: ['facebook', 'twitter', 'linkedin', 'weibo', 'instapaper']
      info: true
always_allow_html: yes
documentclass: article
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

Algebraic perfect hash functions
================================

See the [function reference](https://queelius.github.io/perfect_hash_functions/index.html)

A *set* is an unordered collection of distinct elements.
If we know the elements in a set are given by $a$, $b$, and $c$, we may denote
the set with the notation $\{a,c,b\}$.

A *finite* set has a finite number of elements.
For example, $\{ 1, 3, 5 \}$ is a finite set with three elements.
When sets $\mathbb{A}$ and $\mathbb{B}$ are *isomorphic*, denoted by
$\mathbb{A} \cong \mathbb{B}$, they can be put into a one-to-one correspondence
(bijection), e.g., $\{b,a,c\} \cong \{1,2,3\}$.
Since there exists at least one bijection between isomorphic sets, we can
losslessly convert one to the other and thus, isomorphic sets are in some sense
equivalent.

The cardinality of a finite set $\mathbb{A}$ is the number of elements in the
set, denoted by $|\mathbb{A}|$, e.g., $|\{ 1, 3, 5\}| = 3$.
A *countably infinite* set is isomorphic to the set of natural numbers
$\mathbb{N} = \{1,2,3,\ldots\}$.

Given two elements $a$ and $b$, an ordered pair of $a$ then $b$ is denoted by
$(a,b)$, where $(a,b) = (c,d)$ iff $a = c$ and $b = d$.
Ordered pairs are non-commutative and non-associative, i.e., $(a,b) \neq (b,a)$
if $a \neq b$ and $\bigl(a,(b,c)\bigr) \neq \bigl((b,a),c\bigr)$.

Related to the ordered pair is the Cartesian product.

::: {.definition}
The set
$$
\mathbb{X} \times \mathbb{Y} \coloneqq \left\{(x,y) \colon x \in 
\mathbb{X} \land y \in \mathbb{Y}\right\}
$$
is the Cartesian product of sets $\mathbb{X}$ and $\mathbb{Y}$.
:::
By the non-commutative and non-associative property of ordered pairs, the
Cartesian product is non-commutative and non-associative.
However, they are isomorphic, i.e.,
$\mathbb{X} \times \mathbb{Y} \cong \mathbb{Y} \times \mathbb{X}$.

A *tuple* is a generalization of order pairs which can consist of an
arbitrary number of elements, e.g., $(x_1,x_2,\ldots,x_n)$.

::: {.definition}
The $n$-fold Cartesian product of sets $\mathbb{X_1},\ldots,\mathbb{X_n}$ is
given by
$$
\mathbb{X_1} \times \cdots \times \mathbb{X_n} \coloneqq
  \bigl\{(x_1,\ldots,x_n) : x_1 \in \mathbb{X_1} \land \cdots
  \land x_n \in \mathbb{X_n}\bigr\}.
$$
:::
Note that
$$
\mathbb{X_1} \times \mathbb{X_2} \times \mathbb{X_3} \cong
  \mathbb{X_1} \times \left(\mathbb{X_2} \times \mathbb{X_3}\right) \cong
  \left(\mathbb{X_1} \times \mathbb{X_2}\right) \times \mathbb{X_3},
$$
thus we may implicitly convert between them without ambiguity.

If each set in the $n$-ary Cartesian product is the same, the power notation may
be used, e.g., $\mathbb{X}^3 \equiv \mathbb{X} \times \mathbb{X} \times \mathbb{X}$.
As special cases, the $0$-ary (nullary) Cartesian product is defined to be
$\{ \emptyset \}$ and the $1$-ary (unary) Cartesian product is the identity,
e.g., $\mathbb{X}^1 = \mathbb{X}$.

::: {.definition}
A one-to-one relation between $\{0,1\}^m$ and $\{1,2,\ldots,2^m\}$ is given
by the equation
$$
b_m \Longleftrightarrow \sum_{j=1}^{m} 2^{m - j + 1} b_j\,,
$$
where $b_m \in \{0,1\}^m$.
We denote the transformation in either direction by the unary operation $'$,
e.g., $2' = 0$.
:::

We show some of the mappings in Table \@ref(tab:convert).

Table: (\#tab:convert) A one-to-one relation between $\{0,1\}^*$ and $\mathbb{N}$.


| $\mathbb{N}$ | $\{0,1\}^*$ |
| ------------ | ----------- |  
| 1            | $\emptyset$ |
| 2            | $0_2$       |
| 3            | $1_2$       |
| 4            | $00_2$      |
| 5            | $01_2$      |
| 6            | $10_2$      |
| 7            | $11_2$      |
| 8            | $000_2$     |
| 9            | $001_2$     |
| 10           | $010_2$     |
| 11           | $011_2$     |
| 12           | $100_2$     |
| 13           | $101_2$     |
| 14           | $110_2$     |
| 15           | $111_2$     |
| $\vdots$     | $\vdots$    |


A *hash function* is given by the following definition.

::: {.definition}
Hash functions of type $\mathbb{X} \mapsto \mathbb{Y}$ are just total
functions, normally with a finite codomain, with the weak assumption that they
will be used as a device to assign elements from $\mathbb{X}$ a value from
$\mathbb{Y}$ without requiring any particular rule.
:::
For a given bit string $x$ and hash function $\operatorname{hash}$,
$y = \operatorname{hash}(x)$ is denoted the *hash* of $x$.

We are particularly interested in *perfect* hash functions as given by the
following definition.

::: {.definition}
A perfect hash function over the set $\mathbb{A} \subseteq \mathbb{X}$, denoted
by
$$
\operatorname{hash}_{\mathbb{A}} \colon \mathbb{X} \mapsto \mathbb{Y},
$$
is an injective function when restricted to $\mathbb{A}$.
That is, there are no collisions among elements of $\mathbb{A}$,
$\operatorname{hash}_{\mathbb{A}}(x) \neq \operatorname{hash}_{\mathbb{A}}(y)$
for all $x, y \in \mathbb{A}$, $x \neq y$.
:::
Perfect hash functions are generally *surjective* when not restricted to
the range of elements it perfectly hashes.

The *load factor* is given by the following definition.

::: {.definition name="Load factor"}
\label{def:loadfactor}
The proportion of hashes mapped to by
$\operatorname{hash}_{\mathbb{A}} \colon \mathbb{X} \mapsto \mathbb{N}$ over
subset $\mathbb{A}$ is denoted the load factor.
Specifically, the load factor of $\operatorname{hash}_{\mathbb{A}}$ is a rational number given
by
$$
\frac{|\operatorname{hash}_{\mathbb{A}}(\mathbb{A})|}
  {|{\operatorname{image}(\operatorname{hash}_{\mathbb{A}})}|}.
$$
:::

A *minimal* perfect hash has a load factor $1$.

A perfect hash function of type $\mathbb{X} \mapsto \mathbb{Y}$ over
$\mathbb{A}$ with a load factor $r$ may be denoted by
$\operatorname{hash}_{\mathbb{A}}^{r}$.
If $m = |\{\mathbb{A}\}|$ and we are interested in drawing attention to the
cardinality of the perfect hash function, we may also denote it by
$\operatorname{hash}_{m}^{r}$.


```{r echo=FALSE, ho-plot, out.width='70%', fig.show='hold', fig.cap="This is a figure caption."}
knitr::include_graphics("img/pf.tex")
```

Every hash function in $\mathbb{X} \mapsto \mathbb{Y}$ is a perfect hash
function over some subset of $\mathbb{X}$, e.g., every hash function is
trivially a perfect hash function of the empty set and singleton sets.

Assuming $\mathbb{X}$ and $\mathbb{Y}$ are finite, the set of hash functions of
type $\mathbb{X} \mapsto \mathbb{Y}$, which may also be denoted by
$\mathbb{Y}^{\mathbb{X}}$, has a cardinality
$$
|{\mathbb{Y}}|^{|\mathbb{X}|}.
$$
The set of perfect hash functions over $\mathbb{A} \subseteq \mathbb{X}$ is a
subset of $\mathbb{X} \mapsto \mathbb{Y}$ with the predicate that no collisions
may occur on any pair of elements in $\mathbb{A}$.

